# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

- First we load necessary libraries:

```{r}
# suppress library loading warnings
library(ggplot2, quietly=T, warn.conflicts=F)
library(Hmisc, quietly=T, warn.conflicts=F)
library(dplyr, quietly=T, warn.conflicts=F)
```

- Next we unzip the **activity.zip** file if necessary and then load
**activity.csv** into a dataframe. Note: please change the working
directory to one containing the file on your computer.
```{r}
setwd("~/Documents/coursera/reproducible data/RepData_PeerAssessment1")
if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
df <- csv.get("activity.csv")
```

## Histogram of Total Number of Steps per Day

```{r}
dfm <- df %>% group_by(date)  %>%
    summarise(daily_total = sum(steps, na.rm=T),
              daily_mean = mean(steps, na.rm=T),
              daily_median = median(steps, na.rm=T),
              daily_NA = sum(is.na(steps))
              )

g <- ggplot(dfm, aes(x=date,y=daily_total))
g <- g + geom_bar(stat="identity")
g <- g + geom_bar(width=0.5, fill="white", color="black") + labs(x = "date", y = "number of steps", title = "Total number of steps taken each day") 
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8,
   colour = "grey50"))
g
```

## Daily Means and Medians of Steps Taken
```{r}
dfm %>% select(date, daily_mean, daily_median)
```

## What is the average daily activity pattern?

```{r}
df5 <- df %>% 
    group_by(interval)  %>%
    summarise( interval_mean = mean(steps, na.rm=T))

plot(df5$interval,df5$interval_mean,type = 'l',xlab="interval", ylab="average steps", main="average steps per interval")
max_int <- df5[df5$interval_mean == max(df5$interval_mean),]
abline(v=max_int$interval,h=max_int$interval_mean, col="blue")
```

- The interval with the maximum average steps is `r max_int$interval` with a value of `r round(max_int$interval_mean,1)`.

## Imputing missing values

```{r}


df_av_int <- merge(x=df, y=df5, by="interval", all=T ) %>% 
    mutate(steps = ifelse( is.na(steps), interval_mean, steps)) %>%
    select(- interval_mean)
#df_nona will have the total steps per date after replacing NA step values 
# to the average steps of the interval. 
df_nona <- df_av_int %>%
    group_by(date) %>% 
    summarise(daily_total=sum(steps, na.rm = T),
              daily_mean=mean(steps, na.rm = T),
              daily_median=median(steps, na.rm = T))


#produce graphs
g <- ggplot(df_nona, aes(x=date,y=daily_total))
g <- g + geom_bar(stat="identity")
g <- g + geom_bar(width=0.5, fill="white", color="black") + labs(x = "date", y = "number of steps", title = "Total number of steps taken each day") 
g <- g + theme(axis.text.x = element_text(angle = 90, hjust = 0, size = 7,
   colour = "grey50"))
g

```

- There are a total `r sum(is.na(df$steps))` NA values for the **steps** variable.
- the variable **df_av_int** contains the new data with NA's replaced with 
the mean for the interval.
- With the NA values replaced we get the following means and medians: 

```{r}
df_nona %>% select(date, daily_mean, daily_median)
```

### Observations on the effect of substitutions on the data set

- Given that a day that had any NA's, also had all NA's, or no data at all, one
could not calculate a mean or median for such a day.
- The difference between the data having NA's and the data having substitutions are that
the former data set has more days with 0 steps as the sum of an empty data set is zero. However, in the latter set, the median and mean exist for 
the previously NA days. 
- The numerical difference between the datasets with and without substitions are shown 
below:


```{r}
df_merge = merge(x = dfm, y=df_nona, by="date",all=TRUE)

df_diff <- df_merge %>% 
     select(date, 
            median.NA = daily_median.x, 
            median.sub = daily_median.y,
            mean.NA = daily_mean.x, 
            mean.sub = daily_mean.y ) %>%
    mutate(diff.median = median.NA - median.sub,
            diff.mean = mean.NA - mean.sub)
df_diff
```


## Are there differences in activity patterns between weekdays and weekends?

The following graphics seem to suggest that there is more activity during the morning for weekdays, but more activity throughout the day during the weekend.

```{r}
# Important: use ifelse for vectorized operations !
week_factor <- function ( aDate ) {
    weekday = weekdays(as.Date(aDate))
    ifelse (weekday %in% c("Saturday","Sunday"), "weekend","weekday")
}

df_wd <- df_av_int %>% 
    mutate(weekend = factor(week_factor(date))) %>%
    group_by(weekend,interval) %>% 
    summarise(av_steps=mean(steps, na.rm=T))


qplot( interval, av_steps, data=df_wd, geom="line", facets = weekend ~ ., ylab="steps per interval", main="steps per interval for weekdays and weekend")

```
